---
title: "Benchmarking study"
author: "Alrik Schörling"
date: "9/9/2021"
output: 
  pdf_document: 
    toc: yes
    fig_width: 4
    fig_height: 4
    fig_caption: yes
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: inline
---

# 1. Setup
Set dir
```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
#paths are relative to this R directory (knitr's default root.dir), e.g. data/qrt pcr.csv
```

Check dir
```{r}
getwd()
```

Install packages
```{r}
#if not done, uncomment to install all necessary packages
#install.packages('tidyverse')
#install.packages('lubridate')
#install.packages('broom')
#install.packages('ggplot2')
#install.packages('extrafont')
#install.packages('ggridges')
#install.packages('FactoMineR')
#install.packages('factoextra')
#install.packages('clValid')
#install.packages('ComplexHeatmap')
#install.packages('circlize')
#install.packages('rstatix')
#install.packages('corrr')
#install.packages('ggprism')
#install.packages('onewaytests')
#install.packages('yarrr')
#install.packages('scales')
#install.packages('wesanderson')
#install.packages('viridis')
#install.packages('scico')
#install.packages('BBmisc')
#install.packages('rmarkdown')
#install.packages('pacman')
#install.packages('ggbreak')
#install.packages('effectsize')
#install.packages('coin')
#install.packages('ggcorrplot')
#install.packages('plotly')
#install.packages('tibble')
#install.packages('effectsize')
#install.packages('lme4')
#install.packages('nlme')
#install.packages('ggbreak')
#install.packages('cowplot')

#ochre
#install.packages("devtools")
#devtools::install_github("ropenscilabs/ochRe")

#complex heatmap
#if (!require("BiocManager", quietly = TRUE))
  #install.packages("BiocManager")

#BiocManager::install("ComplexHeatmap")
```

Load packages
```{r}
pacman::p_load(tidyverse, lubridate, broom, #cleaning
               ggplot2, extrafont, ggridges, ggbreak, cowplot, #plotting
               FactoMineR, factoextra, clValid, #multivariate analysis
               ComplexHeatmap, circlize, #heatmaps
               corrr, rstatix, ggprism, onewaytests, effectsize, lme4, nlme, #statistics
               ochRe, yarrr, scales, wesanderson, viridis, scico) #colors
```

Set themes
```{r}
basic_theme <-  theme_bw(base_family = "Helvetica") + 
  theme(panel.border = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        
        axis.ticks = element_line(linewidth = 0.3),
        axis.ticks.length = unit(.1, "cm"), 
        
        axis.text.x = element_text(color = "black", size = 6),
        axis.text.y = element_text(color = "black", size = 6),
        
        axis.title = element_text(size = 6), 
        
        legend.title = element_text(size = 6), 
        legend.text = element_text(size = 6),
        
        strip.text.x = element_text(size = 6),
        strip.text.y = element_text(size = 6), 
        
        strip.background = element_rect(color = NA, fill = NA), 
        plot.title = element_text(color = "black", size = 7))

theme_1 <- basic_theme + theme(axis.line = element_line(linewidth = 0.3))

theme_2 <- basic_theme + 
  theme(axis.line = element_line(linewidth = 0.3), 
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

theme_3 <- basic_theme + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 1), 
        panel.border = element_rect(linewidth = 0.3, fill = NA))

theme_4 <- basic_theme + 
  theme(panel.border = element_rect(linewidth = 0.3, fill = NA))

theme_5 <- basic_theme + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), 
        panel.border = element_rect(linewidth = 0.3, fill = NA))

theme_6 <- theme_2 + 
  theme(strip.text.x = element_text(face = 'italic'))
```

Palettes
```{r}
reef_pal <- ochre_pal(palette = 'healthy_reef')(7)
basel_pal <- unname(piratepal('basel', length.out = 5))

dissoc_pal <- c('#3C358F', '#58BF88', '#FF7769')
flower_pal <- c("#9ca77f", "#4e611f", "#a68750", "#f2cafe", "#a780db")
fruit_punch_pal <- c('#859DD6', '#4D8085', '#5CAB3F', '#FA894C', '#B55E85', '#DBF4B0')

fraction_pal <- c("#0D434F", "#6A5470", "#FFE0C4", "#D5A511", "#AB2A84","#7CB4BF", "#7180C8","#F2C53D", "gray")
```


#2. qRT-PCR quality control
##2.1 Data pt. 1
```{r}
pcr_regional_spec <- read.csv("data/qrt pcr.csv")
pcr_regional_spec
```

Investigate normality 
```{r}
#log10 normalize
pcr_regional_spec_log10 <- pcr_regional_spec |>
  mutate(expr_lvl_log10 = log(expr_lvl)) |>
  arrange(expr_lvl_log10) |>
  filter_all(all_vars(!is.infinite(.))) 

#plot frequency distribution
ggplot(pcr_regional_spec_log10, aes(x = expr_lvl_log10)) + 
  geom_histogram(aes(y =..density..), fill = "lightgray", col = "black") +
  geom_density(col=2, linewidth = 2) + 
  xlab("") + 
  theme_1

#shapiro-wilk test
pcr_regional_spec_log10 |> ungroup() |> shapiro_test(expr_lvl_log10)

#The data is non-normal according to Shapiro-Wilk test but the distribution looks okay. Proceed to plot a heatmap of the expression levels
```

Selecting markers of interest and normalize
```{r}
#define marker of interest
surface_markers <- c("ALCAM", "APCDD1", "IAP", "CNTN2", "CORIN", 
                     "FOLR1", "TPBG", "LRTM1")
regional_markers <- c("FOXG1", "NKX2.1", #forebrain
                      "PAX6", #dorsal
                      "EN1", "LMX1A", "FOXA2", "OTX2", #midbrain
                      "HOXA2") #hindbrain

#subset the dataset
pcr_regional_spec_sel <- pcr_regional_spec |> 
  filter(gene %in% surface_markers | gene %in% regional_markers) |> 
  mutate(gene = factor(gene, 
                       levels = c("APCDD1", "TPBG", "ALCAM", "CORIN", #surface_markers
                                  "IAP", "LRTM1", "CNTN2", "FOLR1", #surface_markers
                                  "FOXG1", "NKX2.1", #forebrain
                                  "EN1", "LMX1A", "FOXA2", "OTX2", #midbrain
                                  "PAX6", #dorsal
                                  "HOXA2"))) |> #hindbrain
  ungroup() 

#summarize the and normalize the dataset
pcr_regional_spec_sel_norm  <- pcr_regional_spec_sel |>
  ungroup() |>
  group_by(gene, region) |>
  summarise(mean_expr_lvl = mean(expr_lvl)) |> 
  ungroup() |> 
  group_by(gene) |> 
  mutate(norm_mean_expr_lvl = BBmisc::normalize(mean_expr_lvl, 
                                                   method = "range")) |> 
  mutate(region = factor(region, 
                         levels = c("dFB", "vFB", "dMB", "dHB", 
                                    "vHB", "rVM", "cVM")))
```

Plot heatmap of regional enrichment
```{r}
plt <- ggplot(pcr_regional_spec_sel_norm, aes(x = region, y = gene, 
                                      fill = norm_mean_expr_lvl)) + 
  geom_tile() +
  scale_fill_scico(palette = "bilbao", direction = -1,
                   name = "Norm. mRNA \n expr. level") + 
  labs(y = "", x = "hESC regional differentiation", title = "Normalized mRNA expression levels") +
  theme_5 + 
  theme(axis.text.y = element_text(face = "italic"),
        legend.key.height=unit(2, "mm"),
        legend.key.width=unit(3, "mm"))

plt
#ggsave("figs/qrt pcr regional spec hmap.pdf", plt, width = 3, height = 2.5, bg = "transparent")
```

Correlating  markers
```{r}
pcr_tbl_vals <- pcr_regional_spec_sel |> 
  pivot_wider(values_from = expr_lvl, names_from = gene) |>
  mutate(region_unique_batch_id_cell_line = paste(region, unique_batch_id, cell_line, 
                                                  sep = "_")) |> 
  relocate(region_unique_batch_id_cell_line) |> 
  select(!c(region, unique_batch_id, cell_line))

pcr_tbl_vals <- column_to_rownames(pcr_tbl_vals, "region_unique_batch_id_cell_line")
pcr_tbl_vals_cor <- (cor(pcr_tbl_vals, method = "pearson"))^2
```

Visualize correlation between markers 
```{r}
#define a color gradient using colorRamp2
sunset_pal <- colorRamp2(breaks = c(0, 1), colors = c('#6BA6DB', '#FFB03B'))
sweden_pal <- colorRamp2(breaks = c(0, 1), colors = c('#006AA7', '#FECC02'))


hmap <- Heatmap(
                #basics
                pcr_tbl_vals_cor, #data
                col = sweden_pal, #colors
                column_title = "Correlation of genes' normalized mRNA expression levels", #column title
                name = "r-squared", #key name
                show_row_dend = F, #logic if remove row dendrogram
                
                #clustering method
                clustering_method_columns = "ward.D2", #clustering cols
                clustering_method_rows = "ward.D2", #clustering rows
                clustering_distance_columns = "pearson", #distance metric
                clustering_distance_rows = "pearson", #distance metric
                column_km = 4, row_km = 3, #no. of clusters
                
                #aesthetics
                row_names_gp = gpar(fontface = "italic", fontsize = 7), 
                row_title_gp = gpar(fontsize = 0), 
                
                column_names_gp = gpar(fontface = "italic", fontsize = 7), 
                column_title_gp = gpar(fontsize = 7), 
                border = TRUE,
                
                heatmap_legend_param = list(
                  title_gp = gpar(fontface = 'plain', fontsize = 7),
                  labels_gp = gpar(fontface = 'plain', fontsize = 7)))

#pdf(file = "figs/qrt pcr corr heatmap.pdf", width = 4, height = 2.5) 
plt <- draw(hmap)
plt
dev.off()
```

Fold change normalized to mean
```{r}
#write function to calculation the sem
sem <- function(x) sd(x)/sqrt(length(x)) #calculate SEM

#normalize data to mean
pcr_regional_spec_sel_norm_mean <- pcr_regional_spec_sel |>
  group_by(gene) |>
  mutate(norm_expr_lvl = expr_lvl/mean(expr_lvl)) |>
  select(-expr_lvl) |>
  ungroup() |>
  group_by(gene, region) |>
  summarise(mean = mean(norm_expr_lvl), sem = sem(norm_expr_lvl)) |>
  ungroup() |>
  group_by(gene) |>
  mutate(mean = 100*(mean/max(mean)), sem = 15 * sem) |> 
  mutate(region = factor(region,
                         levels = c("dFB", "vFB", "dMB", "rVM", "cVM", "dHB",
                                    "vHB")))
```

# Session info
R and package versions used for the analysis (recorded in session_info.txt)
```{r}
sessionInfo()
```

